---
title: "Lab 8 Intro"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("jpeg")
library("tidyverse")
```

## Setup
1. Login to the lab computer.    
2. Navigate to github.com and login.  
3. Find the class repository "datascibiol" (https://github.com/jmledford3115/datascibiol).  
4. Download a copy the class repository to the desktop.  
5. Clone your repository to the desktop using GitHub Desktop.  
6. Copy the files for today's lab from the class repository and paste them into **your** repository.  
7. Open today's lab in RStudio.  

## Warm-up
Use the heartrate data from the previous lab to answer the following questions.  
```{r}
heartrate <- read_csv("data/heartrate.csv")
```

1. Make a barplot that shows the heartrate for drug a for each patient.
```{r}
heartrate %>% 
  select(patient, a) %>%  #use select to get an x and y variable
  ggplot(aes(x =patient, y = a))+ #x is categorical and y is numerical
  geom_col() # could also do geom_bar(stat = "identity")
```

2. Make a barplot that shows the average heartrate for each drug.
```{r}
heartrate %>% 
  pivot_longer(-patient,
               names_to = "drug",
               values_to = "heartrate") %>% 
  group_by(drug) %>% 
  summarise(mean_heartrate = mean(heartrate)) %>% 
  ggplot(aes(x = drug, y = mean_heartrate))+
  geom_col()
```
