---
title: "Homework 9"
author: "Jenna Hussein"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
``` 

## Data
For this homework, we will use a data set compiled by the Office of Environment and Heritage in New South Whales, Australia. It contains the enterococci counts in water samples obtained from Sydney beaches as part of the Beachwatch Water Quality Program. _Enterococci_ are bacteria common in the intestines of mammals; they are rarely present in clean water. So, _Enterococci_ values are a measurement of pollution. `cfu` stands for colony forming units and measures the number of viable bacteria in a sample [cfu](https://en.wikipedia.org/wiki/Colony-forming_unit).   

This homework loosely follows the tutorial of [R Ladies Sydney](https://rladiessydney.org/). If you get stuck, check it out!  

1. Start by loading the data `sydneybeaches`. Do some exploratory analysis to get an idea of the data structure.
```{r}
sydneybeaches <- read_csv("data/sydneybeaches.csv", na = c("NA", " ", ".", "-999", "not measured")) %>% clean_names()
```

2. Are these data "tidy" per the definitions of the tidyverse? How do you know? Are they in wide or long format?
```{r}
glimpse(sydneybeaches)
```
Yes, I think this data is tidy per the definitions of the tidyverse because there is only one piece of data per cell, the variable names are not data, and the data is in long format.

3. We are only interested in the variables site, date, and enterococci_cfu_100ml. Make a new object focused on these variables only. Name the object `sydneybeaches_long`
```{r}
sydneybeaches_long <- sydneybeaches %>% 
  select(site, date, enterococci_cfu_100ml)
```

4. Pivot the data such that the dates are column names and each beach only appears once (wide format). Name the object `sydneybeaches_wide`
```{r}
sydneybeaches_wide <- sydneybeaches_long %>% 
  pivot_wider(names_from = site,
              values_from = enterococci_cfu_100ml)
sydneybeaches_wide
```

5. Pivot the data back so that the dates are data and not column names.
```{r}
sydneybeaches_wide %>% 
  pivot_longer(-date,
               names_to = "site",
               values_to = "enterococci_cfu_100ml")
```

6. We haven't dealt much with dates yet, but separate the date into columns day, month, and year. Do this on the `sydneybeaches_long` data.
```{r}
sydneybeaches_long %>% 
  separate(date, into= c("day", "month", "year"), sep = "/")
```

7. What is the average `enterococci_cfu_100ml` by year for each beach. Think about which data you will use- long or wide.
```{r}
new_long <- sydneybeaches_long %>% 
  separate(date, into= c("day", "month", "year"), sep = "/") %>% 
  select(enterococci_cfu_100ml, year, site) %>% 
  group_by(year, site) %>% 
  summarize(mean_enterococci_cfu_100ml = mean(enterococci_cfu_100ml, na.rm=T, .groups = 'keep')) %>% 
  arrange(-mean_enterococci_cfu_100ml)
new_long
```

8. Make the output from question 7 easier to read by pivoting it to wide format.
```{r}
new_long %>% 
  pivot_wider(names_from = site,
              values_from = mean_enterococci_cfu_100ml)
```

9. What was the most polluted beach in 2013?
```{r}
new_long %>% 
  pivot_wider(names_from = site,
              values_from = mean_enterococci_cfu_100ml) %>% 
  filter(year == 2013)
```
The most polluted beach was Little Bay Beach because it had the highest mean for Enterococci value.


10. Explore the data! Do one analysis of your choice that includes a minimum of three lines of code.

Which site had the least pollution in the month of June?
```{r}
sydneybeaches_long %>% 
  separate(date, into = c("day", "month", "year", sep = "/", na.rm.  = T)) %>% 
  select(site, month, enterococci_cfu_100ml) %>% 
  group_by(site, month) %>% 
  summarize(mean_entero = mean(enterococci_cfu_100ml, na.rm=T)) %>% 
  filter(month == "06") %>% 
  arrange(mean_entero)
```

South Maroubra Beach had the least average pollution in the month of June.

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.   